---
title: "report4"
author: "Viktoria Zruttova"
date: "2024-04-07"
output: html_document
---
```{r}
library(car)
library(caret)
library(ggplot2)
library(dplyr)
library(GGally)
library(readxl)
library(MLmetrics)
```


```{r}
#load the dataset
mydata <- read.csv('./data/day.csv')

```

```{r}
#convert 'dteday' column to Date format
mydata$dteday <- as.Date(mydata$dteday)

#season
mydata$season <- cut(mydata$season,
                     breaks = c(0.5, 1.5, 2.5, 3.5, 4.5),
                     labels = c("Winter", "Spring", "Summer", "Fall"))
mydata$season <- factor(mydata$season, levels = c("Winter", "Spring", "Summer", "Fall"))

#workingday
mydata$workingday <- ifelse(mydata$workingday == 0, "Not_Workingday", "Workingday")
mydata$workingday <- factor(mydata$workingday, levels = c("Not_Workingday", "Workingday"))

#weather
mydata$weathersit <- cut(mydata$weathersit,
                         breaks = c(0.5, 1.5, 2.5, 3.5, 4.5),
                         labels = c("Weather_1", "Weather_2", "Weather_3", "Weather_4"))
mydata$weathersit <- factor(mydata$weathersit, levels = c("Weather_1", "Weather_2", "Weather_3", "Weather_4"))

head(mydata)
```

We are splitting the dataset in 2 parts, first part is training set which is done at year 2011, the second- validation part is done at year 2012.

```{r}
#split the dataset into training and validation sets
training_data <- mydata[mydata$yr == 0, ]
validation_data <- mydata[mydata$yr == 1, ]
```





```{r}
# Set a seed for reproducibility
# Setting a seed is crucial for reproducibility and consistency of results.
# It ensures that when you run the same code multiple times, you'll ...
# ... obtain the same results.
set.seed(20231103)
# Filter dataset for year 2011 (0) and year 2012 (1)
year2011 <- mydata[mydata$yr == 0, ]
year2012 <- mydata[mydata$yr == 1, ]

# Add 'Type' column and assign values
year2011$Type <- "Training"
year2012$Type <- "Validation"

# Combine training and validation data
mydata <- rbind(year2011, year2012)

```

## Boxplots for proposed covariates y

We choose 3 questions of interest:

1. **How does working day affect the number of registered users?** where Y is the numnber of registered users
2. **How does season affect the number of casual users?** where Y is the sqrt(casual users)
3. **Is weather correlated to the number of bike rentals?* ** where Y is the count of bikes rented 

### NOTE - we probably need to do stratified sampling as in lab 9 comment, since we do not have equal representation in both training and validation datasets. We should address this issue - look into this further.

```{r}

#registered
ggplot(mydata, aes(x = Type, y = registered, color = Type)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(x = "Type", y = "# Registered Users") +
  scale_fill_manual(name = "Dataset Type", values = c("Training" = "blue", "Validation" = "red")) +
  ggtitle("# Registered Users by Type (Training vs. Validation)") +
  theme_bw()

#sqrt(casual)
ggplot(mydata, aes(x = Type, y = sqrt(casual), color = Type)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(x = "Type", y = "Square Root # Casual Users") +
  scale_fill_manual(name = "Dataset Type", values = c("Training" = "blue", "Validation" = "red")) +
  ggtitle("Square Root # Casual Users by Type (Training vs. Validation)") +
  theme_bw()

#cnt
ggplot(mydata, aes(x = Type, y = cnt, color = Type)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(x = "Type", y = "# Total Users") +
  scale_fill_manual(name = "Dataset Type", values = c("Training" = "blue", "Validation" = "red")) +
  ggtitle("# Total Users by Type (Training vs. Validation)") +
  theme_bw()

```

```{r}
#creating training data our of mydata (training -> data from 2011)
training_data <- subset(mydata, Type == "Training")
summary(training_data)
```

## Full models for each question
We are fitting full model which will be used to stepwise regression later. We address multicollinearity issue and delete variables that might cause it.

### Note: I am not sure which of the variables should be deleted and which should be kept in the model to make it better.

```{r}


full_model1 <- lm(registered ~ as.factor(workingday) + as.factor(season)+ holiday + weekday + temp + hum + windspeed + as.factor(weathersit),
                 data = training_data)

full_model2 <- lm( sqrt(casual)~ as.factor(workingday) + as.factor(season) + holiday + weekday + temp + hum + windspeed + as.factor(weathersit),
                 data = training_data)

full_model3 <- lm( cnt ~ casual +as.factor(workingday) + as.factor(season) + holiday + weekday + temp + hum + windspeed + as.factor(weathersit),
                 data = training_data)

summary(full_model1)
summary(full_model2)
summary(full_model3)

#plot added variables
avPlots(full_model1)
avPlots(full_model2)
avPlots(full_model3)
```


# Stepwise Regression

```{r}
stepwise_model1 <- step(full_model1, direction = "both")
stepwise_model2 <- step(full_model2, direction = "both")
stepwise_model3 <- step(full_model3, direction = "both")
```

```{r}
summary(stepwise_model1)
summary(stepwise_model2)
summary(stepwise_model3)
```

```{r}
avPlots(stepwise_model1)
avPlots(stepwise_model2)
avPlots(stepwise_model3)
```

For vif, values close to 1 indicate no multicollinearity, values between 1-2 indicate mild collinearity which is not ideal but can be considered ok since it won't severaly impact the interpretability of the coefficients, values aboved 2 indicate severe multicollinearity which is considered unreliable.

```{r}
vif(stepwise_model1)
vif(stepwise_model2)
vif(stepwise_model3)
```

### Prediction


```{r}
validation_data$predicted_registered <- predict(stepwise_model1, newdata = validation_data)
validation_data$predicted_sqrtcasual <- predict(stepwise_model2, newdata = validation_data)
validation_data$predicted_count <- predict(stepwise_model3, newdata = validation_data)
```


### Note: here in the predicitons I am unsure why the firts one is negative, in thesecond quetsion, I am unsure whether I should square the original value or not (not sure wthere it predicts the squared casaul count or not). In the third question, it produces Nans???

```{r}
observed_values1 <- validation_data$registered
predicted_values1 <- validation_data$predicted_registered
# Calculate different prediction performance metrics
# Functions from the MLmetrics package
# Common regression metrics
# Calculate the Root Mean Squared Error (RMSE), which measures the ...
# ... average magnitude of prediction errors.
# Lower is better.
rmse1 <- RMSE(predicted_values1, observed_values1)
# same as
# sqrt(mean((observed_values - predicted_values)^2))

# Compute the Mean Absolute Error (MAE), indicating the average absolute ...
# ... difference between predicted and observed values.
# Lower is better.
mae1 <- MAE(predicted_values1, observed_values1)
# same as
# mean(abs(observed_values - predicted_values))

# Calculate the Mean Absolute Percentage Error (MAPE), measuring the ...
# ... average percentage difference between predicted and observed values.
mape1 <- MAPE(predicted_values1, observed_values1)
# same as
# mean(abs(predicted_values-observed_values)/observed_values)

# Determine the R-squared (R²) Score, representing the proportion of the ...
# ... variance in the observed values (of validation data set) ... 
# ... explained by the predicted values from the model.
# Higher is better.
r_squared1 <- R2_Score(predicted_values1, observed_values1)
# same as
# summary(lm(observed_values ~ predicted_values))$r.squared

# Display the calculated metrics
cat("Root Mean Squared Error (RMSE):", round(rmse1, digits = 4), "\n")
cat("Mean Absolute Error (MAE):", round(mae1, digits = 4), "\n")
cat("R-squared (R^2) Score:", round(r_squared1, digits = 4), "\n")
cat("Mean Absolute Percentage Error (MPE):", round(mape1, digits = 4), "\n")



#second question

observed_values2 <- sqrt(validation_data$casual) #should I transform it here??
predicted_values2 <- validation_data$predicted_sqrtcasual
rmse2 <- RMSE(predicted_values2, observed_values2)
mae2 <- MAE(predicted_values2, observed_values2)
mape2 <- MAPE(predicted_values2, observed_values2)
r_squared2 <- R2_Score(predicted_values2, observed_values2)
cat("Root Mean Squared Error (RMSE):", round(rmse2, digits = 4), "\n")
cat("Mean Absolute Error (MAE):", round(mae2, digits = 4), "\n")
cat("R-squared (R^2) Score:", round(r_squared2, digits = 4), "\n")
cat("Mean Absolute Percentage Error (MPE):", round(mape2, digits = 4), "\n")



#third question
observed_values3 <- validation_data$count
predicted_values3 <- validation_data$predicted_count
rmse3 <- RMSE(predicted_values3, observed_values3)
mae3 <- MAE(predicted_values3, observed_values3)
mape3 <- MAPE(predicted_values3, observed_values3)
r_squared3 <- R2_Score(predicted_values3, observed_values3)
cat("Root Mean Squared Error (RMSE):", round(rmse3, digits = 4), "\n")
cat("Mean Absolute Error (MAE):", round(mae3, digits = 4), "\n")
cat("R-squared (R^2) Score:", round(r_squared3, digits = 4), "\n")
cat("Mean Absolute Percentage Error (MPE):", round(mape3, digits = 4), "\n")


```


## Diagnostics

```{r}

```

